---
Title: Exploratory Data Analysis of Instacart
Output:
  html_notebook: default
---


#Import the dataset 
```{r}
Aisles = read.csv("/Users/shimonyagrawal/Desktop/Instacart-market-basket-analysis/aisles.csv")
Departments = read.csv("/Users/shimonyagrawal/Desktop/Instacart-market-basket-analysis/departments.csv")
Order_Prior = read.csv("/Users/shimonyagrawal/Desktop/Instacart-market-basket-analysis/order_products__prior.csv")
Orders = read.csv("/Users/shimonyagrawal/Desktop/Instacart-market-basket-analysis/orders.csv")
Products = read.csv("/Users/shimonyagrawal/Desktop/Instacart-market-basket-analysis/products.csv")
```
#install packages
```{r}
install.packages("DBI")
install.packages("odbc")
install.packages("tidyverse")
install.packages("lubridate")
install.packages("GGally")
install.packages("forecast")
install.packages("ggplot2")
install.packages("readr")
install.packages("dplyr")
install.packages("treemap")
install.packages("scales")
install.packages("tidyr")
install.packages("arules")
install.packages("arulesViz")
install.packages("methods")
install.packages("plyr")

library(DBI)
library(odbc)
library(tidyverse)
library(lubridate)
library (GGally)
library(forecast)
library(ggplot2)
library(readr)
library(dplyr)
library(treemap)
library(scales)
library(tidyr)
library(arules)
library(arulesViz)
library(methods)
library(plyr)
```

#Create a main data frame of the data set with relevant information 
```{r}
eda_df <- Products %>% 
  left_join(Aisles, by = "aisle_id") %>% 
  left_join(Departments, by = "department_id") %>% 
  left_join(Order_Prior, by = "product_id") %>%
  left_join(Orders, by = "order_id") %>% 
  drop_na()

view(eda_df)

```

#How many items do people order in a single purchase? 
```{r}
Order_Prior %>% 
  group_by(order_id) %>% 
  dplyr::summarize(n_items = last(add_to_cart_order)) %>%
  ggplot(aes(x=n_items)) + 
  geom_histogram(stat = "count", fill = "khaki") +
  geom_rug() +
  coord_cartesian(xlim = c(0,60))
  
```

#What are Instacart's most ordered items?
```{r}
MostSold = eda_df %>%
  select(product_id, product_name) %>%
  group_by(product_name) %>%
  dplyr::summarise(TotalOrders = n())

Top15Sold <- MostSold %>%
  filter(TotalOrders > 92957)

ggplot(data=Top15Sold, aes(x = product_name, y = TotalOrders)) +
  geom_bar(stat = "identity", fill = "khaki2") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
```

#What products do people order the most: organic vs non organic products?
```{r}
Organic_Nonorganic <- Products %>%
  mutate(organic=ifelse(str_detect(str_to_lower(Products$product_name),'organic'),"organic","not organic"), organic= as.factor(organic))

OrganicPro <- Organic_Nonorganic %>%
  filter(organic == "organic") %>%
  left_join(eda_df, by = "product_id") %>%
  select (product_id, product_name.x) %>% 
  group_by(product_name.x) %>%
  dplyr::summarise(TotalOrganicProducts = n())

Top15OrganicProducts <- OrganicPro %>%
  filter(TotalOrganicProducts > 78805)

ggplot(data=Top15OrganicProducts, aes(x=product_name.x, y=TotalOrganicProducts))+
  geom_bar(stat="identity", fill="lightpink")+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=1))+
  ggtitle("Top 15 Organic Products")

NonorganicPro <- Organic_Nonorganic %>%
  filter(organic == "not organic") %>%
  left_join(eda_df, by = "product_id") %>%
  select (product_id, product_name.x) %>% 
  group_by(product_name.x) %>%
  dplyr::summarise(TotalNonOrganicProducts = n())

Top15NonorganicProducts <- NonorganicPro %>%
  filter(TotalNonOrganicProducts > 56768)

ggplot(data=Top15NonorganicProducts, aes(x=product_name.x, y=TotalNonOrganicProducts))+
  geom_bar(stat="identity", fill="slategray2")+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=1))+
  ggtitle("Top 15 Non-Organic Products")
  
```

#What hour of day and day of week most products are ordered?
```{r}
#hour of day
OrderByHour <- Orders %>%
  select(order_id, order_hour_of_day)%>%
  group_by(order_hour_of_day)%>%
  dplyr::count()

ggplot(data=OrderByHour, aes(x=order_hour_of_day, y= n))+
  geom_line(color = "palevioletred") +
  ggtitle("Products ordered by Hour of Day") +
  xlab("Hours of Day") +
  ylab("No. of Products")
  
MeanOrderHour <- Orders %>%
  dplyr::summarise(AvgHour = mean(order_hour_of_day))
  
#day of week
OrderByDay <- Orders %>%
  select (order_id, order_dow) %>%
  group_by(order_dow) %>%
  dplyr::count ()

ggplot(data=OrderByDay, aes(x=order_dow, y=n))+
  geom_line(color = "palevioletred") +
  ggtitle ("Product Ordered by Days of Week") +
  xlab("Days of Week") +
  ylab ("No. of Products")

MeanOrderWeek <- Orders %>%
  dplyr::summarise(AvgDay = mean(order_dow))
  
```

#What items are first added to the cart by Instacart's users? 
```{r}
priority <- Order_Prior %>% 
  group_by(product_id, add_to_cart_order) %>% 
  dplyr::summarize(count = n()) %>%
  mutate(average=count/sum(count)) %>%
  filter(add_to_cart_order==1, count>10) %>%
  arrange(desc(average)) %>%
  mutate(average1= average*100000) %>%
  left_join(Products, by="product_id") %>%
  select(product_name, average1, average, count) %>%
  ungroup() %>%
  top_n(10, wt=average1)

products <- c("Emergency Contraceptive", "Energy Iced Tea" ,"California Champagne",
              "Cabernet Sauvignon", "Flavoured Vodka", "Draft Sake", "Organic Rasberry Tea",
              "Soy Powder Infant Formula", "Nasal Decongestant Inhaler", "Infant Formula")

count <- c(37,51,14,14,50,25,27,24,30,29)

ggplot(data = priority,
       aes(x = c("Emergency Contraceptive", "Energy Iced Tea" ,"California Champagne",
                 "Cabernet Sauvignon", "Flavoured Vodka", "Draft Sake", "Organic Rasberry Tea",
                 "Soy Powder Infant Formula", "Nasal Decongestant Inhaler", "Infant Formula"),
           y= average1)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=1))+
  ggtitle("Top Priority Products") +
  xlab ("Product Names") +
  ylab ("Frequency")
  
```

#What are Instacart's most reordered products? 
```{r}
MostReordered_Number<- Order_Prior%>%
  group_by(product_id)%>%
  dplyr::summarize(proportion_reordered = mean(reordered), n=n())%>%
  top_n(10, wt=n)%>%
  arrange(desc(proportion_reordered))%>%
  left_join(Products,by="product_id")

#pie chart
names <- c("Banana", "Bag of Organic Bananas", "Organic Whole Milk", "Organic Hass Avocado",
           "Organic Strawberries", "Organic Baby Spinach", "Organic Avocado",
           "Strawberries", "Larger Lemon", "Limes")
n <- c(472565, 379450, 137905,213584,264683,	241921,176815,142951,152657,140627)

pie(n, names, main = "Most Reordered Products")

```
#When do Instacart's users place the next order? 
```{r}
Orders %>% 
  ggplot(aes(x=days_since_prior_order)) + 
  geom_histogram(stat="count",fill="lightblue2")
  
```
